---
title: "Linear Optimization"
format: html
editor: visual
---

# Linear Optimization

## Libraries Needed

Load these libraries into your environment.

```{r}
# Install the lpSolve package if you haven't
#install.packages("lpSolve")

# Load the package
library(lpSolve)


```

# Min Example

## Step 1 --- Mathematical Formulation of the Linear Programming Model

### **Decision Variables**

Let

\[

```{=tex}
\begin{aligned}
A &= \text{number of gallons of product A to produce} \\
B &= \text{number of gallons of product B to produce}
\end{aligned}
```
\]

------------------------------------------------------------------------

### **Objective Function**

Minimize the total production cost:

$$
Z = 2A + 3B 
$$

------------------------------------------------------------------------

### **Subject to the Constraints**

1.  **Customer demand for product A:**\
   $$A \ge 125 $$

2.  **Minimum combined production requirement:**\
    $$ A + B \ge 350 $$

3.  **Processing time limitation (600 hours):**\
    $$ 2A + B \le 600 $$

4.  **Non-negativity constraints:**\
    $$A,B \ge 0 $$

This code will walk you through the set up of the minimization progment and then the constraints

### Minimization

```{r}

# Objective coefficients (for 2A + 3B)
objective <- c(2, 3)

```
### Constraints
```{r}
# Constraint matrix
constraints <- matrix(c(
  1, 0,   # A >= 125
  1, 1,   # A + B >= 350
  2, 1    # 2A + B <= 600
), nrow = 3, byrow = TRUE)

# Direction of constraints
direction <- c(">=", ">=", "<=")

# Right-hand side of constraints
rhs <- c(125, 350, 600)
```
### Linear Formula

```{r}
# Solve the LP
solution <- lp(direction = "min", objective.in = objective, const.mat = constraints, const.dir = direction,const.rhs = rhs)

```

### Answer

```{r}

# Display results
solution$objval       # Minimum total cost
solution$solution      # Optimal values for A and B
```

# Max Example

## Step 1 --- Welte Mutual Funds

### **Decision Variables**

Let

\[

```{=tex}
\begin{aligned}
x_1 &= \text{dollars invested in Atlantic Oil} \\
x_2 &= \text{dollars invested in Pacific Oil} \\
x_3 &= \text{dollars invested in Midwest Steel} \\
x_4 &= \text{dollars invested in Huber Steel} \\
x_5 &= \text{dollars invested in Government Bonds}
\end{aligned}
```
\]

------------------------------------------------------------------------

### **Objective Function**

Maximize the total projected return:

$$ \text{Maximize } Z = 0.073x_1 + 0.103x_2 + 0.064x_3 + 0.075x_4 + 0.045x_5 $$

------------------------------------------------------------------------

### **Subject to the Constraints**

1.  **Total investment (available funds):**\
    $$  x_1 + x_2 + x_3 + x_4 + x_5 = 100{,}000$$ 

2.  **Oil industry limit:**\
    $$  x_1 + x_2 \le 50{,}000 $$ \

3.  **Steel industry limit:**\
        $$   x_3 + x_4 \le 50{,}000 $$ 

4.  **Government bonds requirement (≥ 25 % of steel investment):**\
        $$   x_5 \ge 0.25(x_3 + x_4)     $$   or equivalently
        $$  -x_5 + 0.25x_3 + 0.25x_4 \le 0     $$  
        

5.  **Pacific Oil limit (≤ 60 % of total oil investment):**\
        $$   x_2 \le 0.60(x_1 + x_2)     $$  
        or equivalently
        $$   -0.6x_1 + 0.4x_2 \le 0     $$  

6.  **Non-negativity:**\
        $$   x_1,,x_2,,x_3,,x_4,,x_5 \ge 0     $$  
        
### Maximizations        

```{r}
# Objective coefficients
objective <- c(0.073, 0.103, 0.064, 0.075, 0.045)
```

Constraints

```{r}
# Constraint matrix
constraints <- matrix(c(
  # x1  x2   x3   x4   x5
   1,   1,   1,   1,   1,      # total investment = 100000
   1,   1,   0,   0,   0,      # oil ≤ 50000
   0,   0,   1,   1,   0,      # steel ≤ 50000
   0,   0,   0.25, 0.25, -1,   # bond ≥ 0.25*(steel)
  -0.6, 0.4,  0,   0,   0      # Pacific Oil ≤ 0.6*(oil)
), nrow = 5, byrow = TRUE)

# Direction of constraints
directions <- c("=", "<=", "<=", "<=", "<=")

# Right-hand side of constraints
rhs <- c(100000, 50000, 50000, 0, 0)
```

### Linear Formula
```{r}
# Solve the LP
solution <- lp(direction = "max",
               objective.in = objective,
               const.mat = constraints,
               const.dir = directions,
               const.rhs = rhs)
```

### Answer
```{r}
# Results
solution$status        # 0 = optimal
solution$objval        # Maximum return
solution$solution      # Optimal investment in x1-x5
```
